Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-05-04 Paper_Draft2/Draft2.Rmd
Logistic regression model

The AIF360 toolkit contains various group and individual fairness metrics as well as pre-processing, in-processing, and post-processing algorithms that we used to debias the COMPAS algorithm [@aif360-oct-2018]. Fairness metrics are mathematical measures of whether an algorithm treats members of different groups (such as racial or gender groups) equally. An algorithm has no understanding of the historical oppression of certain groups and how such bias is baked into the data on which the algorithm is trained. Thus, fairness metrics provide a method of evaluating an algorithm’s level of bias towards or against the unprivileged group versus the privileged group. We researched the definitions and applications of different fairness metrics [@ashokan2021fairness] to determine which metric would be most appropriate for our project.

@@

Pre-processing refers to mitigating bias within the training data, and it is the most flexible method because it has not yet trained a model that may carry assumptions about the data. It is important to keep in mind that pre-processing prevents assumptions in the modeling, but does not account for the bias in data collection. Training data is where bias is most likely to be introduced. We use the reweighing pre-processing algorithm from AIF360 which assigns weights to the data. “The advantage of this approach is, instead of modifying the labels, it assigns different weights to the examples based upon their categories of protected attribute and outcome such that bias is removed from the training dataset. The weights are based on frequency counts. However as this technique is designed to work only with classifiers that can handle row-level weights, this may limit your modeling options” [@Ronaghan2019AI]. We used a logistic regression model for this algorithm, as it is the easiest to interpret in the given context. After running the fairness metrics using the pre-processing algorithm, we were able to compare our results to the baseline metrics from the previous section.

>>>>

The AIF360 toolkit contains various group and individual fairness metrics as well as pre-processing, in-processing, and post-processing algorithms that we used to debias the COMPAS algorithm [@aif360-oct-2018]. Fairness metrics are mathematical measures of whether an algorithm treats members of different groups (such as racial or gender groups) equally. An algorithm has no understanding of the historical oppression of certain groups and how such bias is baked into the data on which the algorithm is trained. Thus, fairness metrics provide a method of evaluating an algorithm’s level of bias towards or against the unprivileged group versus the privileged group. We researched the definitions and applications of different fairness metrics [@ashokan2021fairness] to determine which metric would be most appropriate for our project. To begin, we implemented a logistic regression model predicting recidivism using the defendant data. This analyzes the baseline values without any bias mitigation. We use logistic regression, as it is the easiest model to interpret in the given context.

@@

Pre-processing refers to mitigating bias within the training data, and it is the most flexible method because it has not yet trained a model that may carry assumptions about the data. It is important to keep in mind that pre-processing prevents assumptions in the modeling, but does not account for the bias in data collection. Training data is where bias is most likely to be introduced. We use the reweighing pre-processing algorithm from AIF360 which assigns weights to the data. “The advantage of this approach is, instead of modifying the labels, it assigns different weights to the examples based upon their categories of protected attribute and outcome such that bias is removed from the training dataset. The weights are based on frequency counts. However as this technique is designed to work only with classifiers that can handle row-level weights, this may limit your modeling options” [@Ronaghan2019AI]. After running the fairness metrics using the pre-processing algorithm, we were able to compare our results to the baseline metrics from the previous section.
